```{r, error=TRUE, results='hide', message=FALSE}
#BiocManager::install("edgeR")
#BiocManager::install("Glimma")

library(readxl)
library("edgeR")
library(Glimma)
library(dplyr)
library("data.table")
library(DESeq2)
library(ggplot2)
#library("glmpca")

```




Wczytanie plików
```{r, error=TRUE, results='hide', message=FALSE}
getwd()
dir.create(file.path(getwd(), "results"))

#coldata <- read_excel(file.path(getwd(),"input_data/coldata_all.xlsx"))
coldata <- read.table(file.path(getwd(),"input_data/coldata.csv"), header = TRUE, sep=",")
coldata$names <- coldata$Condition

counts <- read.table(file.path(getwd(),"input_data/Counts_with_names.csv"), header = TRUE, sep=",")

results <- read.table(file.path(getwd(),"input_data/Results.txt"), header=TRUE)
results_Y <-  results %>% filter(MIRNA == "Y")
results_Y$known_miRNAs <- ifelse(is.na(results_Y$known_miRNAs), results_Y$MajorRNA, results_Y$known_miRNAs)
results_Y$known_miRNAs <- ave(as.character(results_Y$known_miRNAs), results_Y$known_miRNAs, FUN=function(x) if (length(x) > 1) paste0( x[1], '(', seq_along(x), ')') else x[1])
results_Y_short <- results_Y[,c("Locus", "Strand", "MajorRNA", "known_miRNAs")]

results <- results[,c("Name", "known_miRNAs"  , "MajorRNA")]


raw_counts <- left_join(counts, results, by = "Name")

 
raw_counts_filtered <- raw_counts %>% filter(MIRNA == "Y")
raw_counts_filtered$known_miRNAs <- ifelse(is.na(raw_counts_filtered$known_miRNAs), raw_counts_filtered$MajorRNA, raw_counts_filtered$known_miRNAs)

raw_counts_filtered$known_miRNAs <- ave(as.character(raw_counts_filtered$known_miRNAs), raw_counts_filtered$known_miRNAs, FUN=function(x) if (length(x)>1) paste0(x[1], '(', seq_along(x), ')') else x[1])

raw_counts_named <- raw_counts_filtered[, !names(raw_counts_filtered) %in% c("Coords","Name", "MIRNA", "known_miRNAs", "MajorRNA")]
#raw_counts_named <- raw_counts_filtered[,c("Hgy3_apex_rep1","Hgy3_apex_rep2","Hgy3_apex_rep3", "Hgy3_1_2_rep1", "Hgy3_1_2_rep2","Hgy3_1_2_rep3", "Hgy3_3.5_rep1", "Hgy3_3.5_rep2", "Hgy3_3.5_rep3", "Hgy3_6_8_rep1", "Hgy3_6_8_rep2", "Hgy3_6_8_rep3", "Hgy3_L_rep1", "Hgy3_L_rep2", "Hgy3_L_rep3")]

#colnames(raw_counts_named) <- c("Hgy3_apex_rep1","Hgy3_apex_rep2","Hgy3_apex_rep3", "Hgy3_1_2_rep1", "Hgy3_1_2_rep2","Hgy3_1_2_rep3", "Hgy3_3_5_rep1", "Hgy3_3_5_rep2", "Hgy3_3_5_rep3", "Hgy3_6_8_rep1", "Hgy3_6_8_rep2", "Hgy3_6_8_rep3", "Hgy3_L_rep1", "Hgy3_L_rep2", "Hgy3_L_rep3")
                                           
rownames(raw_counts_named) <- as.vector(raw_counts_filtered$known_miRNAs)




write.csv(file.path(getwd(),"results/counts_filtered.csv"))
write.csv(file.path(getwd(),"results/counts_filtered_named.csv"))



row_names <- rownames(raw_counts_named)


```




Ustalenie Group do design
```{r, error=TRUE, results='hide', message=FALSE}
#Group <- factor(paste(coldata$sex_type_of_plant,coldata$sex_type_of_flower,coldata$Line ,sep="."))
Group <- factor(coldata$Condition)
```


Etap filtracji - filterByExpr - Filter genes by expression level. Determine which genes have sufficiently large counts to be retained in a statistical analysis.
```{r, error=TRUE}


y <- DGEList(raw_counts_named, group=Group)
rownames(y) <- row_names




keep <- filterByExpr(y, group=Group)

y_filter <- y[keep, , keep.lib.sizes=FALSE]
salmon.g.filtered  <- y[keep, keep.lib.sizes=FALSE]
cat("Przed filtracją:", length(y$counts), "Po filtracji:", length(salmon.g.filtered$counts))

```



```{r, error=TRUE, results='hide', message=FALSE}

dge <- DGEList(salmon.g.filtered, group = Group)
dge <- calcNormFactors(dge)

normalized.counts <- cpm(dge)
log_normalized.counts <- cpm(dge, log=TRUE)

write.csv(normalized.counts, file.path(getwd(),"results/CPM_counts.csv"))
write.csv(log_normalized.counts, file.path(getwd(),"results/log_CPM_counts.csv"))

```


Exploring differences between libraries
```{r, error=TRUE}

pch <- c(0,1,2,15,16,17, 18, 19 , 20)
colors <- rep(c("darkgreen", "red", "blue", "green", "orange"), 2)
plotMDS(y, col=colors[Group], pch=pch[Group])
legend("top", legend=levels(Group), pch=pch, col=colors, ncol=2)

plotMDS(y, col=colors[Group], pch=pch[Group])

```

```{r, error=TRUE, results='hide', message=FALSE}
 #mod1  <- model.matrix(~0+ condition, coldata)
 mod1  <- model.matrix(~0+ Group)
 design <- mod1
```


Wizualna inspekcja
```{r, error=TRUE}

y <- dge
y <- estimateDisp(y,design, robust=TRUE)
plotBCV(y)
fit <- glmQLFit(y,design, robust=TRUE)
head(fit$coefficients)
plotQLDisp(fit)
summary(fit$df.prior)
```




Exploring differences between libraries
```{r, error=TRUE}

pch <- c(0,1,2,15,16,17, 18, 19 , 20)
colors <- rep(c("darkgreen", "red", "blue", "green", "orange"), 2)
plotMDS(y, col=colors[Group], pch=pch[Group])
legend("bottom", legend=levels(Group), pch=pch, col=colors, ncol=2)

plotMDS(y, col=colors[Group], pch=pch[Group])

```
```{r, error=TRUE}
jpeg(file=file.path(getwd(),"results/MDS_plot.jpg"))
pch <- c(0,1,2,15,16,17, 18, 19 , 20)
colors <- rep(c("darkgreen", "red", "blue", "green", "orange"), 2)
plotMDS(y, col=colors[Group], pch=pch[Group])
legend("bottom", legend=levels(Group), pch=pch, col=colors, ncol=2)
dev.off()


plotMDS(y, col=colors[Group], pch=pch[Group])

```


```{r, error=TRUE}

glMDSPlot(y, labels=paste(coldata$Sample_name ), 
          groups=coldata$Condition, launch=TRUE)

glimmaMDS(y, labels=paste(coldata$Sample_name ), 
          groups=coldata$Condition, launch=TRUE, html=file.path(getwd(),"results/MDS_plot.html"))
unlink(paste0(getwd(),"/results/", "MDS_plot_files"), recursive = TRUE)

```




```{r, error=TRUE}
for (i in 1:ncol(y)){
  plotMD(y, column=i)
  abline(h=0, col="red", lty=2, lwd=2)
}

```

Utworzenie kontrastów:
```{r, error=TRUE, results='hide', message=FALSE}

#Utworzenie macierzy kontrastów: Grupa_badawcza vs kontrola

my.contrasts <- makeContrasts(
  Line_Hgy3_1_2vsLine_Hgy3_apex = GroupHgy3_1_2 - GroupHgy3_apex ,
  Line_Hgy3_3_5vsLine_Hgy3_1_2 = GroupHgy3_3_5 - GroupHgy3_1_2 ,
  Line_Hgy3_6_8vsLine_Hgy3_3_5 = GroupHgy3_6_8 - GroupHgy3_3_5 ,
  Line_Hgy3_1_2vsLine_Hgy3_L = GroupHgy3_1_2 - GroupHgy3_L ,
  Line_Hgy3_3_5vsLine_Hgy3_L = GroupHgy3_3_5 - GroupHgy3_L ,
  Line_Hgy3_6_8vsLine_Hgy3_L = GroupHgy3_6_8 - GroupHgy3_L 
  
  
  
  
  
  , levels = design 
)

contrasts_names <- c("Line_Hgy3_1_2vsLine_Hgy3_apex",  "Line_Hgy3_3_5vsLine_Hgy3_1_2", "Line_Hgy3_6_8vsLine_Hgy3_3_5", "Line_Hgy3_1_2vsLine_Hgy3_L" , "Line_Hgy3_3_5vsLine_Hgy3_L", "Line_Hgy3_6_8vsLine_Hgy3_L")




```




```{r, error=TRUE}
qlf_test <- function(con, contr_name) {
  dir.create( paste0(getwd(),"/results/results_",contr_name), showWarnings = TRUE)
  dir.create(paste0(getwd(),"/results/results_",contr_name, "/qlf_test"), showWarnings = FALSE)
  qlf <- glmQLFTest(fit, contrast=con,)
  res_qlf <- as.data.frame(qlf)
  res_qlf$miRNA <- rownames(res_qlf)
  #colnames(res_qlf) <- c("logFC", "logCPM", "F", "PValue", "FDR")
  colnames(results_Y_short)[colnames(results_Y_short) == "known_miRNAs"] <- "miRNA"
  res_qlf <- left_join(res_qlf, results_Y_short, by="miRNA")
  res_qlf <- res_qlf[,c("miRNA", "Locus", "Strand", "MajorRNA", "logFC", "logCPM", "F", "PValue")]
  write.csv(res_qlf, file = file.path(paste0(getwd(),"/results/results_",contr_name, "/qlf_test"), "res_qlf.csv"))
  
  top_tags <- as.data.frame(topTags(qlf))
  top_tags$miRNA <- rownames(top_tags)
  top_tags <- left_join(top_tags, results_Y_short, by="miRNA")
  top_tags <- top_tags[,c("miRNA", "Locus", "Strand", "MajorRNA", "logFC", "logCPM", "F", "PValue", "FDR")]
  write.csv(top_tags, file = file.path(paste0(getwd(),"/results/results_",contr_name, "/qlf_test"), "qlf_topDEgenes.csv"))
  
  is.de <- decideTests(qlf)
  is.de05 <- decideTests(qlf, p.value=0.05)
  
  capture.output(summary(is.de), file = file.path(paste0(getwd(),"/results/results_",contr_name, "/qlf_test"), "qlf_isde.txt"))
  capture.output(summary(is.de05), file = file.path(paste0(getwd(),"/results/results_",contr_name, "/qlf_test"), "qlf_isde05.txt"))

  pdf(file.path(paste0(getwd(),"/results/results_",contr_name, "/qlf_test"), "MD_plot.pdf"))
  plotMD(qlf, status=is.de)
  dev.off()
  return(qlf)
}
```


```{r, error=TRUE}
LRT_test <- function(con, contr_name){
  dir.create(paste0(getwd(),"/results/results_",contr_name), showWarnings = FALSE)
  dir.create(paste0(getwd(),"/results/results_",contr_name, "/lrt_test"), showWarnings = FALSE)
  fit_LRT <- glmFit(y,design)
  lrt <- glmLRT(fit_LRT,contrast=con)
  res_lrt <- as.data.frame(lrt)[1:100,]
  write.csv(res_lrt, file = file.path(paste0(getwd(),"/results/results_",contr_name, "/lrt_test"), "res_lrt.csv"))
  
  top_tags <- as.data.frame(topTags(lrt, n=100))
  write.csv(top_tags, file = file.path(paste0(getwd(),"/results/results_",contr_name, "/lrt_test"), "lrt_topDEgenes.csv"))
  
  
  o <- order(lrt$table$PValue)
  capture.output(summary(de <- decideTests(lrt)), file = file.path(paste0(getwd(),"/results/results_",contr_name, "/lrt_test"), "lrt_isde.txt"))
  print(summary(de <- decideTests(lrt)))
  detags <- rownames(y)[as.logical(de)]
  
  pdf(file.path(paste0(getwd(),"/results/results_",contr_name, "/lrt_test"), "lrt_plot.pdf"))
  plotSmear(lrt, de.tags=detags)
  abline(h=c(-1, 1), col="blue")
  dev.off()
  return(lrt)
}

```


```{r, error=TRUE} 
#Differential expression above a fold-change threshold
treat_test <- function(con, contr_name){
  dir.create(paste0(getwd(),"/results/results_",contr_name), showWarnings = FALSE)
  dir.create(paste0(getwd(),"/results/results_",contr_name, "/treat_test"), showWarnings = FALSE)
  tr <- glmTreat(fit, contrast=con, lfc=log2(1.5))
  res_tr <- as.data.frame(tr)[1:100,]
  write.csv(res_tr, file = file.path(paste0(getwd(),"/results/results_",contr_name, "/treat_test"), "res_treat.csv"))
  
  print(topTags(tr))
  top_tags <- as.data.frame(topTags(tr, n=100))
  write.csv(top_tags, file = file.path(paste0(getwd(),"/results/results_",contr_name, "/treat_test"), "treat_topDEgenes.csv"))
  
  
  is.de <- decideTests(tr)
  capture.output(summary(is.de), file = file.path(paste0(getwd(),"/results/results_",contr_name, "/treat_test"), "treat_isde.txt"))
  print(summary(is.de))
  
  pdf(file.path(paste0(getwd(),"/results/results_",contr_name, "/treat_test"), "MD_plot.pdf"))
  plotMD(tr, status=is.de)
  dev.off()
  return(tr)
  }

```

```{r, error=TRUE}
#a heatmap to visualize the top 30 DE genes according to the test
heat_map <- function(contr_name, y, st_test, folder_name){
  dir.create(paste0(getwd(),"/results/results_",contr_name, "/", folder_name), showWarnings = FALSE)
  logCPM <- cpm(y, prior.count=2, log=TRUE)
  rownames(logCPM) <- rownames(y$counts)
  colnames(logCPM) <- paste(y$samples$group, 1:3, sep="-")
  o <- order(st_test$table$PValue)
  logCPM <- logCPM[o[1:30],]
  
  pdf(file.path(paste0(getwd(),"/results/results_",contr_name,"/", folder_name), "heatmap.pdf"))
  coolmap(logCPM, margins=c(7,7), lhei=c(1,6), lwid=c(1,3))
  dev.off()
}
```



Funkcja do analizy DE
```{r, error=TRUE}
DE_analyse <- function(con, contr_name, y){
  
  qlf <- qlf_test(con, contr_name)
  try(lrt <- LRT_test(con, contr_name))
  tr <- treat_test(con, contr_name)
# 
  
  glimmaVolcano(qlf, group=y$samples$group,  counts = y$counts, html = paste0(getwd(),"/results/results_", contr_name, "/", contr_name, "_volcano.html"), main = paste0(contr_name, "_Volcano_plot") )

  glimmaMA(qlf, group=y$samples$group,  counts = y$counts, html =  paste0(getwd(),"/results/results_", contr_name, "/", contr_name, "_MA_plot.html"), main = paste0(contr_name, "_MA_plot" ))
  unlink(paste0(getwd(),"/results/results_", contr_name, "/", contr_name, "_MA_plot_files"), recursive = TRUE)
  unlink(paste0(getwd(),"/results/results_", contr_name, "/", contr_name, "_volcano_files"), recursive = TRUE)
  
  
  heat_map(contr_name, y,  tr, "treat_test")
  heat_map(contr_name, y,  qlf, "qlf_test")
  heat_map(contr_name, y,  lrt, "lrt_test")

}

```




Pętla porównująca kontrasty DE
```{r, error=TRUE}
for (i in 1:length(contrasts_names)){
#for (i in 1:3){
  print(i)
  con <- my.contrasts[,contrasts_names[i]]
  DE_analyse(con, contrasts_names[i], y)
  
}

```


```{r}

  dds_to_pca <- DESeq2::DESeqDataSetFromMatrix(countData=raw_counts_named, colData=coldata, design=~Condition)
  vsd <- rlog(dds_to_pca)
  keep <- filterByExpr(dds_to_pca, group=Group)
 # keep <- rowSums(counts(dds_to_pca)) >= 3
  dds_to_pca <- dds_to_pca[keep,]


  pcaData <- plotPCA(vsd, intgroup = c( "Condition"), returnData = TRUE)#zmienić tutaj
  pcaData
  percentVar <- round(100 * attr(pcaData, "percentVar"))


  pca_plot1 <- ggplot(pcaData, aes(x = PC1, y = PC2, color = Condition)) +#i tutaj zmienić kolor
    geom_point(size =3) +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    coord_fixed() +
    ggtitle("PCA with rlog data")

  ggsave(paste0(getwd(),"/results/", "PCA_plot_all_samples.jpg"), pca_plot1, width = 10, height = 10, units = "cm", dpi = 300,)
  ggsave(paste0(getwd(),"/results/",  "/PCA_plot.pdf"), pca_plot1, width = 10, height = 10, units = "cm", device = "pdf")




  # gpca <- glmpca(counts(dds_to_pca), L=2)
  # gpca.dat <- gpca$factors
  # gpca.dat$name <- dds_to_pca$name
  # gpca.dat$condition <- dds_to_pca$condition
  # pca_plot2 <- ggplot(gpca.dat, aes(x = dim1, y = dim2, color = vsd$Sample_name, shape = vsd$Condition)) +
  #   geom_point(size =3) + coord_fixed() + ggtitle("glmpca - Generalized PCA")
  # 
  # ggsave(paste0(getwd(),"/results/", "/PCA_plot_all_samples_glmpca.jpg"), pca_plot2)
  # ggsave(paste0(getwd(),"/results/",  "/PCA_plot_glmpca.pdf"), pca_plot2, width = 20, height = 20, units = "cm", device = "pdf")
  # 
  # 




```



